---
title: "Biostat 203B Homework 1"
author: Zian ZHUANG
subtitle: Due Jan 22 @ 11:59PM
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->
<style type="text/css">
body{ /* Normal  */
      font-size: 17px;
      font-family: "Times New Roman";
  }
h1,h2,h3,h4,h5,h6{
  font-family: "Times New Roman";
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!-- Begin writing -->

Display machine information for reproducibility:
```{r}
sessionInfo()
```

## Q1. Git/GitHub
**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

### Q1.1
Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

> **solution**: Done.

### Q1.2
Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

> **solution**: Done.

### Q1.3
Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

> **solution**: Done.

### Q1.4
After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

> **solution**: Done.

### Q1.5
After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

> **solution**: Sounds very nice. Thanks!

## Q2. Linux Shell Commands

### Q2.1
This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

> **solution**: The verification links to the [completion report](https://www.citiprogram.org/verify/?k600f5f26-0a64-4f74-a92f-83c92ec84f0e-40386397) and [completion certificate](https://www.citiprogram.org/verify/?w4f427623-e63f-402b-b4df-48fd01dd09a6-40386397). 

### Q2.2
The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

> **solution**: Done.

### Q2.3
Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

> **solution**:
    ```{bash}
    ls -l /usr/203b-data/mimic-iv/core
    ```
    ```{bash}
    ls -l /usr/203b-data/mimic-iv/hosp
    ```
    ```{bash}
    ls -l /usr/203b-data/mimic-iv/icu
    ```

> The functionalities of bash commands:

> - `zcat`: Line utility for viewing the contents of a compressed file without literally uncompressing it.
- `zless`: A filter which allows examination of compressed or plain text files one screenful at a time on a soft-copy terminal.
- `zmore`: A file perusal filter for crt viewing of compressed text.
- `zgrep`: Search out expressions from a given a file even if it is compressed.
 
### Q2.4
What's the output of following bash script?
    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
> **solution**: The bash script print out all `.gz` files in the folder `core`.

Display the number of lines in each data file using a similar loop.

> **solution**:

    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
        echo "the number of lines:" 
        zcat $datafile | awk 'END { print NR }'
      done
    ```

### Q2.5
Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

> **solution**:
```{bash}
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk '(NR<=5)'
```
```{bash}
echo "the number of rows:"
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk 'END { print NR }' 
```
```{bash}
echo "the number of unique patients:"
zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | 
awk  -F ',' '{ print $1 }' | uniq | wc -l
```

## Q3. Who's popular in Price and Prejudice

### Q3.1
You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

> **solution**: Use `grep` prints lines that match an name and then calculated the times.
```{bash}
echo "number of times Elizabeth. is mentioned:"
grep 'Elizabeth' pride_and_prejudice.txt | wc -l
echo "number of times Jane is mentioned:"
grep 'Jane' pride_and_prejudice.txt | wc -l
echo "number of times Lydia is mentioned:"
grep 'Lydia' pride_and_prejudice.txt | wc -l
echo "number of times Darcy is mentioned:"
grep 'Darcy' pride_and_prejudice.txt | wc -l
```

### Q3.2 
What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

> **solution**: `'> test1.txt'` redirects output to `test1.txt`, overwriting the file. `'>> test1.txt'`redirects output to `test1.txt`, appending the redirected output at the end.

### Q3.3 
Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

> **solution**: 
```{bash}
./middle.sh pride_and_prejudice.txt 20 5
```
> the meaning of:

> - `"$1"`: the first column of the input (the element `pride_and_prejudice.txt` here)
- `"$2"`: the second column of the input (the element `20` here)
- `"$3"`: the third column of the input (the element `5` here)

> The first line `#!/bin/sh` means that the script should always be run with bash, rather than another shell. It's a convention for the server to know what program it should use to run the shell script.

## Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

> **solution**: 
```{bash}
cal
```
> `cal` display the calender of current month.
```{bash}
cal 2021
```
> `cal 2021` display the calender of all month in 2021.
```{bash}
cal 9 1752
```
> `cal 9 1752` seems there some days missing in September 1752. The Gregorian calendar reform was adopted by the Kingdom of Great Britain in September 1752. As a result, the September 1752 cal shows the adjusted days missing. [[wiki](https://en.wikipedia.org/wiki/Cal_(Unix))]
```{bash}
date
```
> `date` returns the date in the default system timezone. 
```{bash}
hostname
```
> `hostname` provides the name of the server.
```{bash}
arch
```
> `arch` provides the computer architecture.
```{bash}
uname -a
```
> `uname -a` prints the name, version and other details about the current machine and the operating system running on it.
```{bash}
uptime
```
> `uptime` returns information about how long your system has been running together with the current time, number of users with running sessions, and the system load averages for the past 1, 5, and 15 minutes.
```{bash}
who am i
```
> `who am i` displays the username of the current user when this command is invoked.
```{bash}
who
```
> `who` displays account information: user login name, user's terminal, time of login as well as the host the user is logged in from.
```{bash}
w
```
> `w` displays information about currently logged in users and what each user is doing.
```{bash}
id
```
> `id` print real and effective User ID (UID) and Group ID (GID).
```{bash}
last | head
```
> `last | head` displays the first 10 users logged in and out since the file /var/log/wtmp was created.
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```
> `echo {con,pre}{sent,fer}{s,ed}` generates all the permutations possible of a set of elements ({con,pre}{sent,fer}{s,ed}) stored in a variable in groups of 2 elements.
```{bash}
time sleep 5
```
> `time sleep 5` pauses execution of shell scripts or commands for a 5-second period on a Linux 
```{bash}
history | tail
```
> `history | tail` shows 10 of the last commands that have been recently used.
