---
title: "Biostat 203B Homework 3"
author: Zian ZHUANG
subtitle: Due Mar 5 @ 11:59PM
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->
<style type="text/css">
body{ /* Normal  */
      font-size: 17px;
      font-family: "Times New Roman";
  }
h1,h2,h3,h4,h5,h6{
  font-family: "Times New Roman";
}
</style>
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(DBI)
library(dbplyr)
library(RPostgreSQL)
library(tidyverse)
library(lubridate)
library(data.table)
library(magrittr)
library(tidyverse)
```
<!-- Begin writing -->

Display machine information for reproducibility:
```{r}
sessionInfo()
```

## Q1. Compile the ICU cohort in HW2 Q8 from the PostgreSQL database `mimiciv`. 
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
is_teaching_server = Sys.info()[["nodename"]] == "ucla-biostat-203b"
```

Below is an outline of steps.

### Q1.1 
Connect to database `mimiciv`. We are going to use username `postgres` with password `postgres` to access the `mimiciv` database.
```{r}
# Load configuration settings
#host  <- '127.0.0.1'
#port  <- '5432'
user  <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b", 
                'postgres', # on teaching server
                'huazhou')  # on my MacBook
password <- ifelse(Sys.info()[["nodename"]] == "ucla-biostat-203b",
                   'postgres', # on teaching server
                   'huazhou')  # on my MacBook   
dbname <- 'mimiciv'
# Connect to the database using the configuration settings
(con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                  dbname = dbname, 
                  user = user, 
                  password = password))
```

### Q1.2
List all schemas in the `mimiciv` database.
```{r}
dbGetQuery(con, "SELECT SCHEMA_NAME FROM INFORMATION_SCHEMA.SCHEMATA")
```

List tables in the `mimiciv` database:
```{r}
dbListTables(con)
```

List schemas and tables in the `mimiciv` database (bash command).
```{bash, include = is_teaching_server, eval = is_teaching_server}
psql -U postgres -d mimiciv -c "\dt *."
```

### Q1.3
Connect to the icustays table. Note how to use `Id()` to specify the schema containing the table.
```{r}
icustays_tble <- tbl(con, Id(schema = "mimic_icu", table = "icustays")) %>%
  print(width = Inf)
```

### Q1.4
Connect to the patients table.
```{r}
patients_tble <- tbl(con, Id(schema = "mimic_core", table = "patients")) %>%
  print(width = Inf)
```

### Q1.5
Connect to the admissions table.
```{r}
admissions_tble <- tbl(con, Id(schema = "mimic_core", table = "admissions")) %>%
  print(width = Inf)
```

### Q1.6
Connect to the `mimic_labevents_icu` table.
```{r}
labevents_tble <- tbl(con, Id(schema = "public", 
                              table = "mimic_labevents_icu")) %>%
  print(width = Inf)
```

### Q1.7
Connect to `mimic_chartevents_icu` table.
```{r}
chartevents_tble <- tbl(con, Id(schema = "public", 
                              table = "mimic_chartevents_icu")) %>%
  print(width = Inf)
```

### Q1.8
Put things together. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) keep only the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) display the SQL query, (vi) collect SQL query result into memory as a tibble, (vii) create an indicator for 30-day mortality, (viii) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
} 
```

```{r}
icustays_tble %>%
  # keep only the first ICU stay of each unique patient
  group_by(subject_id) %>% 
  filter(rank(intime) == 1) %>% 
  ungroup() %>% 
  # merge in admissions and patients tables
  left_join(patients_tble, by = c("subject_id")) %>%
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>%
  # keep adults only (age at admission >= 18)
  mutate(age_at_adm = year(admittime) - anchor_year + anchor_age) %>%
  filter(age_at_adm >= 18) %>%
  # merge in the labevents and chartevents tables
  left_join(labevents_tble, by = c("subject_id", "hadm_id")) %>%
  left_join(chartevents_tble, by = c("subject_id", "hadm_id")) %>%
  # display the SQL query
  show_query() %>%
  # collect SQL query result into memory as a tibble
  collect() %>% 
  # create an indicator for 30-day mortality
  mutate(death_30 = ifelse((deathtime - admittime)/(60*24) <= 30,
                           "Yes", "No")) %>%
  mutate_at(vars(death_30), 
            function(x){ifelse(is.na(x) == TRUE, "No", x)}) %>%
  # save the final tibble
  saveRDS(file = "./mimiciv_shiny/icu_cohort.rds")
```

Close database connection and clear workspace.
```{r}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>